# (PART) R Language {-} 

# Objects

> To understand computations in R, two slogans are helpful:
>
> * Everything that exists is an object.
> * Everything that happens is a function call.
>
> --- John Chambers


## Inspect object
* `str()`: struction of object, get unsystematic overview
* `typeof()`: "low-level" type of an object
* `mode()`: storage mode of an object
* `class()`: "high-level" class of object
    - Determines what method is applied when passed to generic functions

`typeof()` and `mode()` are almost the same (at the same level) except that they use different names

| `typeof()`      | `mode()`           |
|-----------------|--------------------|
|`integer`, `double`|`numeric`         |
|`special`, `builtin`|`function`       |
|`symbol`         |`name`              | 
|`language`       |`(`, `call`         |

`class()` is on a complete different level and determines what method is applied when passed to generic functions. Some classes are explicitly stated as an attribute, others are implicitly given by the `mode`

For historical reasons (back compatibility with S / S+), another low-level inspector function is `storage.mode()`.


```{r, echo = FALSE}
l.ex <- list(one = 1, s1 = 1:2, pi = pi, I = 1i, `NA` = NA, let = c("A", "a", "b", ":::"),
             fn = mean, fn2 = c, fnSpec = `function`,
             n = as.name("Mä"), n2 = quote(x), nb = quote(`{`),
             ex = expression(1+1), cl = call("round",10), cl2 = quote(sin(x)),
             formula = y ~ x+t, formals = formals(lm), arg1 = formals(lm)[[1]],
             NULL = NULL)
# str(l.ex, vec.len = 16, max.level = 1)

myShow <- function(x, max.length= 1000L) {
    r <- tryCatch(format(x), error=function(e)e)
    r <- if(inherits(r, "error"))
             tryCatch(as.character(x), error=function(e)e)
         else paste(r, collapse = " ")
    r <- if(inherits(r, "error"))
             tryCatch(capture.output(x), error=function(e)e)
         else paste(r, collapse = " ")
    substr(r, 1L, max.length)
}

cbind(show    = sapply(l.ex, myShow, max.length = 16),
      typeof  = sapply(l.ex, typeof),
      storage.mode = sapply(l.ex, storage.mode),
      mode    = sapply(l.ex, mode),
      class   = sapply(l.ex, class)) -> tab.ex
knitr::kable(tab.ex, format="pandoc")
```



## Hirarchy

Objects can be ordered in a hirarchical structure

* Vector
    + Atomic vector
        - Matrix
        - Array
    + List
        - Data frame
* Function
    + Primitive
    + Closure
* Environment
* Symbol
* Call
* Expression
* Forumla
* ...

Some objects are building blocks for other objects

* Symbols can be combined to calls.
* Calls can be combined to expressions


## Basic Objects

### Vectors, functions and environments
Vectors will be covered in chapter \@ref(vec), functions in  chapter \@ref(fun) and environments in chapter \@ref(env)



### Symbols
* A way to refer to R objects by name
* Constituent of calls
* Type: symbol
* Class: name

```{r}
x <- call("sin", 1:3)
typeof(x[[1]])
```



### Calls
* An unevaluated expression which consists of the named function applied to the given arguments
* Class: call
* Type: language
* Evaluate a call with `eval()`

| Action      | R command                            |
|-------------|--------------------------------------|
| Create      | `call()` or `quote()`|
| Check       | `is.call()`                          |
| Coerce      | `as.call()`                          |

```{r}
x <- call("sin", 1:3)
y <- quote(sin(1:3))
eval(x)
```



### Expression
* Lists of calls
* `expression()` returns its arguments as a vector of unevaluated expressions.
* `quote()` returns its argument as an unevaluated expression.
* `eval()` evaluates an expression in a certain environment

```{r}
eval(quote(mean(c(1:3))))
```



### Not understand yet
* `parse()`
* `all.equal()`


Why?
```{r}
x <- call("sin", 1:10)
y <- quote(sin(1:10))
identical(x,y)
```

