# Functions {#fun}

## Types {#FunTypes}

Find function "type" with `typeof()`

* builtin (primitive functions e.g. `sum`)
* special (primitive functions e.g. `[`)
* closure (functions built upon primitive functions, they always have their own environment)

### Primitive functions

* Implemented in C
* Separated into builtin and special
    - builtin: arguments evaluated before passed to C code
    - special: unevaluated arguments passed to C code
* C code called via `.Primitive()` or `.Internal()`
    - `.Primitive()` for pure C code
    - `.Internal()` for mixtures of R and C code
* `body()`, `formals()` and `environment()` are `NULL` (see \@ref(function-components))
* All in base package
* Check with `is.primitive()`

Many primitive function can be used without the classical syntax functionname() (see \@ref(special-calls)). Backticks ` ` ` let you refere to objects that have otherwise reserved or illegal name.





## Function components {#function-components}

+---------------------+----------------------------------------------------------------------------+----------------------------------+
| - `body()`          | Code inside the function.                                                  |                                  |
+---------------------+----------------------------------------------------------------------------+----------------------------------+
| - `formals()`       | List of formal **arguments** which controls how you can call the function. | Chapter \@ref(function-arguments)   |
+---------------------+----------------------------------------------------------------------------+----------------------------------+
| - `environment()`   | “Map” of the location of the function’s variables.                         | Chapter \@ref(function-environment) |
+---------------------+----------------------------------------------------------------------------+----------------------------------+

Like other objects, functions have certain attributes

* `srcref` (source reference): Attribute which all functions have. Returns source code (similar to `body()` but with comments and formatting)
* additional attributes possible

```{r, collapse=TRUE}
myfun <- function(x) {
  x^2 # with comment
  }
attributes(myfun)
body(myfun)
```

## Arguments {#function-arguments}

* Formal arguments (stated in function definition)
* Actual arguments (explicetly passed to the function when you call it)

* Get formal argument: `formals()`
* Set formal arguments: `` `formals<-` ``  (replacement function: change default formals, add formals, remove formals)
* `formals()` without arguments will list the arguments of the function which calls it

```{r, collapse=TRUE}
myfunction <- function(x = 1, y = 2) x + y
formals(myfunction)
myfunction()
formals(myfunction) <- list(x = 3, y = 4)
myfunction()
(function(x) formals())()
```

Arguments are matched by the following hirarchy

* Complete name
* Partial name (prefix maching)
* Position


Use `do.call()` to pass arguments containd in a list to a function.

Use `missing()` to check if a formal argument has been manually set.

### Lazy evaluation:

* Arguments are evaluated when they are used
* If they are never used within a function, they might never be evaluated $\Rightarrow$ not saved in the environment of that function
* To avoid problems in **loops which return functions**, make sure to `force()` the evaluation of the objects which are needed for the returned function
* Example:
    - `add()` is a function in the global environment which returns a function
    - Each function in the list `adders_bad`/`adders_good` are in separate environments
    - without `force(x)` `x` is **not** evaluated in those environments
    - `x` is lazily evaluated the first time that you call one of the adder functions. At this point, the loop is complete and the final value of `x` is 10. Therefore `x` is evaluated in all environments to have a value of 10.


```{r, collapse=TRUE}
# wrong
add <- function(x) function(y) x + y
adders_bad <- vector("list", 10)
for (i in seq_along(adders_bad)) adders_bad[[i]] <- add(i)
adders_bad[[1]](10) # should be 11
adders_bad[[10]](10) ## should be 20
get("x", envir = environment(adders_bad[[1]])) # should be 1

# good
add <- function(x){
  force(x)
  function(y) x + y
}
adders_good <- vector("list", 10)
for (i in seq_along(adders_good)) adders_good[[i]] <- add(i)
adders_good[[1]](10) # should be 11
adders_good[[10]](10) ## should be 20
get("x", envir = environment(adders_good[[1]])) # should be 1
```


* An unevaluated argument is called a **promise**. A promise is made up of two parts:
    - The expression which gives rise to the delayed computation.
    - The environment where the expression was created and where it should be evaluated.

### Elipsis `...`

This argument will match any arguments not otherwise matched, and can be easily passed on to other functions.


## Environment {#function-environment}

### Lexical scoping

* Scoping: Where to look for objects
* Lexical: Scoping is determined where the function is defined (not where it is called)

Rules

* If a name is not defined inside a function, R will look one level up
* Example: x is searched in the environment of each function in the following order h $\rightarrow$ g $\rightarrow$ f $\rightarrow$ global environment

```{r}
x <- 2
f <- function(){
  g <- function() {
    h <- function() x
    h()
  }
  g()
}
f()
```

* Assigning a function also assignes a certain environment (environment property)
* Example:
    - Function `j` returns a function (which connects `x` and `y`)
    - Assigning `j(1)` to `k` makes `k` an function with the environment property pointing to the environment created by running `j()` (the function where the output function was defined). This environment **contains x and y**

```{r}
j <- function(x) {
  y <- 2
  function() {
    c(x, y)
  }
}
k <- j(1)
k()
```

* Object saved in a higher environment are not found below
* Example:
    * `a` is not saved except within function
    * A new call of `j` will not find previous `a`
    * Functon will always return 1 exept if we save `a`

```{r}
j <- function() {
  if (!exists("a")) {
    a <- 1
  } else {
    a <- a + 1
  }
  a
}
j()
```



## Return value

* The last expression evaluated in a function becomes the return value
* `return()` usful to stop evaluation and return error
* `invisible()` usful to invisible return an object (e.g. for ploting functions)
* pure functions: functions which change nothing except for what they return
* Few non-pure functions
    - `library()` which loads a package, and hence modifies the search path.
    - `setwd()`, `Sys.setenv()`, `Sys.setlocale()` which change the working directory,
    - environment variables, and the locale, respectively.
    - `plot()` and friends which produce graphical output.
    - `write()`, `write.csv()`, `saveRDS()`, etc. which save output to disk.
    - `options()` and `par()` which modify global settings.
    - S4 related functions which modify global tables of classes and methods.
    - Random number generators which produce different numbers each time you run them.
* `on.exit()` evaluates expression when function call is finished
* Example

```{r}
in_dir <- function(dir, code) {
  old <- setwd(dir)
  on.exit(setwd(old))
  
  ## what function does...
}
```






## Special calls {#special-calls}

| name                   | description                                  | Built-in examples                               | Your own   |
|------------------------|----------------------------------------------|-------------------------------------------------|------------|
| Control flow operators | Decide which line is evaluated               | `if`, `else`, `for`, `while`, `break`           |            |
| Subsetting operators   | Subset an object                             | `[`, `[[`, `$`                                  |            |
| Infix operators        | function name comes in between ist arguments | `%in%`, `%*%`, `+`, `*`, `^`, `>`, `&`, `<-`, … | `%myname%` |
| Replacement functions  | Replace an existing object                   | `<-`, `[<-`, `names<-`, `levels<-`, …           | `myname<-` |

All special functions can also be called like a "normal" function

```{r}
`[`(letters[1:10], 2)
```

Replacement function make some changes to the data and assign the new output to the first argument

```{r}
x <- 1:10
names(x) <- letters[1:10] # user friendly
x <- `names<-`(x, letters[1:10]) ## behind scenes this is executed
```


You can create your own infix and replacement functions. This is a useful example of providing a default value in case the output of another function is `NULL`:

```{r, eval=FALSE}
`%||%` <- function(a, b) if (!is.null(a)) a else b
function_that_might_return_null() %||% default value
```



## Generic functions

* Depending on the class of the first argument, different **methods** of a function are called by `UseMethod()`
* All available methods of a functionname can be shown with `methods()`.
* methods ending with an asterisks `*` are not exported in the namespace, meaning they should not explicitly be called by the user.
    - get their source code with `getAnywhere()`
```{r}
print
head(methods(print))
```

