# Data handling

## Atomic vectors


### Basic operations

|  | Arithmetic |   | Comparison |   | Logical |
|:-|:-----------|:--|:-----------|:--|:--------|
|`+`| Addition | `<` | less than | `!x` | logical NOT |
|`-`| Subtraction | `>` | greater than | `x & y` | logical AND (vectorized)|
|`*`| Multiplication | `<=` | less or equal | `x && y` | logical AND (single value)
|`/`| Division | `>=` | greater or equal | `x | y` | logical OR (vectoriyed) |
|`^`| Power of | `==` | equal | `x || y` | logical OR (single value) |
|`%%`| Modulo | `!=` | unequal | `xor(x,y)` | logical exclusive OR |
|`%/%`| Integer division | | | | |


### Factors

Also see chapter \@ref(vec-factors)

* `levels()` print or overwrite labels of factor levels
* `relevel()` The levels of a factor are re-ordered so that the level specified by `ref` is first and the others are moved down.
* `reorder()` Reorder the factor levels according to another variable

```{r, out.width='50%'}
# Make level B the first level (reference value)
InsectSprays$spray <- relevel(InsectSprays$spray, ref = "B")

# Reorder levels of spray by median of count (first class should have lowest count median)
InsectSprays$spray <- reorder(InsectSprays$spray, InsectSprays$count, FUN = median)
boxplot(count ~ spray, data = InsectSprays)
```


##### forcats package {-}

* `fct_reorder()`  Reorder the factor levels according to another variable
* `fct_reorder2()`  Reorder the factors according to last value of x axis
* `fct_inorder()`  Reorder factors levels by first appearance or frequency
* `fct_relevel()`  Set a certain level first or last (or at certain position)
* `fct_infreq()`  Reorder levels by the frequency in which they appear in the data (highest frequency first)
* `fct_rev()`     Reverse order of factor levels
* `fct_lump()`    Combine some factor levels to new factor level

```{r}

```


<a href="figures/03_pdf/lubridate.pdf" target="_blank">Test</a>




### Time


### Strings

| Function | Action |
|:----------------|:------------------------------|
| `nchar(x)` | Number of letters |
| `gsub('Peter', 'Pete', x)` | Replaces all Peter with Pete |
| `sub('Peter', 'Pete', x)` | Replaces first Peter with Pete |
| `grep('Pete', x)` | Which entry in a vector contain 'Pete' (returns rownumber) |
| `strsplit(x, '_')` | Splits strings at '_' and makes list out of them |
| `substr(x, 1, 1)` | Extract or replace substrings in a character vector (from start to stop, here: first letter) |
| `match()` | Returns a vector of the positions of (first) matches |
| `%in%` | Similar to match, returns a logical vector |
| `toupper(x)` | Makes all letters uppercase |
| `tolower(x)` | Makes all letters lowercase |



## Lists

### Apply familiy

### Long and wide format

### sQL joins