# (PART) R Programming {-} 

# Basics

## Objects in Enviornment

```{r, eval=FALSE}
ls()      # list all objects in environment
rm()			# remove specific objects
rm(list=ls())                  # remove all Objects
rm(list=setdiff(ls(), "keep")) # keep only one object
```


## Functions

* `function()`  Define function
* `...`         Ellipsis, placeholder for some arguments within a function
* `stop()`			interrupts function
* `print()`			print object (Within functions, R is not autoprinting)
* `return()`		specify explicit what to return (otherwise last object called)
* `invisible()`	returns invisible (just for assignment `<-`)
* `missing()`		returns `TRUE` if an argument has not been specified explicitly. 

##### Examples {-}

```{r, eval=FALSE}
FunNew <- function(x = 1, y = "lalala", ...){
  if(x > 10){
    stop("`x` bigger than 10")  # function stoped with error
  }
  print(y)
  return(x+1)
  x+2
}
```


### Function debugging

* `traceback()` stack of functions called before error occured
* `browser()` enters browser mode if called within function
    - Within brower mode, you can interact with objects in function environment
    - Hit Q => exit browser mode
    - Hit Enter => evaluate next line of function
    - In RStudio
        + Browser mode starts automatically if error occures
        + Set Breakpoints/`browser()` by clicking in script panel on code number (signaled by red dot)


## Control structures

### Conditional statements

```{r, eval=FALSE}
## if else
if (condition) {
  # do that
} else if (condition2) {
  # do something else
} else {
  # 
}

## Vectorized
ifelse(condition, yes, no)
```

For condition use `TRUE`, `FALSE`, `>`, `<`, `is.na()`, `!`, `%in%`, `||`, `&&`, ...

as soon as `||` sees the first `TRUE` it returns `TRUE` without computing anything else. As soon as `&&` sees the first `FALSE` it returns `FALSE`.


```{r, eval=FALSE}
# cut() splits vector into factor groups
# instead of: if(x < -3) {x <- “low”} else …
x <- rnorm(100)
cut(x, breaks = -3:3)

# switch() jumpes to certain operation (e.g. plus, minus, …)
switch_fun <- function(x, y, op) {
  switch(op,
         plus = x + y,
         minus = x - y,
         stop("Unknown op!")
  )
}
```


### Loops

##### Types of loops {-}

* `for(i in 1:10) {}` for-loop, most useful
* `while(cond) {}`		while-loop, most general
* `repeat {}`     		repeat-loop, needs somewhere  `if(cond){break}` to leave loop

##### Components {-}

1. **Output** Before you start the loop, you have to allocate sufficient space for the output
    - `output <- vector("list", length(x))`
    - instead of `list` use `"integer"`, `"double"`, `"character"`, `"logical"`
    - instead of `length(x)` use `nrow()`, `ncol()`
2. **Sequence** Loop over numeric indices
    - `i in seq_along(x)`
    - Alternative
        + loop over names		`(i in names(x))`
        + loop over elements	`(i in x)`
        + problematic for NA and not unique names/elements
3. **Body**
    - `output[[i]] <- my_function(x[[i]])`


##### Preallocation {-}


```{r, eval=FALSE}
## Preallocation
out <- vector(mode = "integer", length = 10)
for(i in 1:10){
  out[i] <- i^2
}

## Never use growing output (slow)
Output <- NULL
for(i in 1:10){
  x <- i^2
  Output <- c(Output, x)
}
```


Unknown output length

* Save it in a list and vectorise it later 		`unlist()`
* Save it in a list and combine it later 		`cbind()`, `rbind()`
* Save it in a character vector and paste it later	`str_c()`

```{r, eval=FALSE}
means <- c(1,50,100)
out <- vector("list", length(means))
for (i in seq_along(means)) {
  n <- sample(100, 1)
  out[[i]] <- rnorm(n, means[[i]])
}
unlist(out)

```














