# Performance

* `system.time()`
    - returns the CPU time used by an expression
    - works well for relatively long computation times, not well for very short computation times
    - runs the expr only once
* `proc.time()`
    - similar to `system.time()`
* `microbenchmark::microbenchmark()`
    - works well for very small timing steps
    - runs the expr 100 times and calculates average runtimes

## In general

1) Look for existing solutions.
2) Do less work.
3) Vectorise.
4) Parallelise.
5) Avoid copies.
6) Byte-code compile


## Profiling

Example:

1) Open an Rprof file and start profiling
2) Run a process
3) Stop profiling
4) Analyze the functions run during profiling

```{r, eval=FALSE}
Rprof("dumkoll.Rprof", interval = 0.005) # start profiling
dd <- dumkoll(50000, df=TRUE)
Rprof(NULL) # stop profiling
sr <- summaryRprof("dumkoll.Rprof")
sr
```



## Microbenchmark

Compare small pieces of code (here function1, function2, function3)

```{r, eval = FALSE}
library(microbenchmark)
mbd <- microbenchmark(function1(n), function2(n), function3(n))
```



## Use package compiler


First compile a function before you apply it to data

```{r, eval = FALSE}
library(compiler)
dumkoll2 <- cmpfun(dumkoll)
dumkoll2(<data>)
```



## Parallelise

Useful packages:

- foreach
- doParallel

Example:

```{r}
library(doParallel)
detectCores() # How many cores are available on my machine
```


