# Functional programming

* Functionals: functions that take a function as an argument
* Closures: functions written by another function
* Anonymous function: function without name


## Closures {#FunProgClosures}
* Not the same meaning as type `closure` (Chapter \@ref(FunTypes))
* Functions which were written by another function
    - Have non-trivial environments (not `R_GlobalEnv` or a package's environment)
* Enclosing environment is the function environment of the function which reated them
    - The function "Enclose" this non-trivial environment
    - Can access all its variables
* Two levels of parameters
    - Child level (Function environment): does the work
    - Parent level (Environment property): control parameter

```{r}
power <- function(exponent) {
  function(x) x ^ exponent
}

square <- power(2)
square(2)
```

Note: function environment of the call `power(2)` is not deleted in this case because it is the enclosing environment of the `square()` function.


## Lists of functions

Useful to apply several functions to one object

```{r}
x <- 1:10
funs <- list(
  sum = sum,
  mean = mean,
  median = median
)
lapply(funs, function(f) f(x))
```

