# Remote Servers

## UZH Thinlinc

* Make IMATH account https://www.math.uzh.ch/li/index.php?id=1794
* Install Thinlinc client https://www.math.uzh.ch/li/?thinlinc
* Setup SSH keys https://wiki.math.uzh.ch/public/ssh



## ETH Euler

https://scicomp.ethz.ch/wiki/Getting_started_with_clusters

## Renkulab

https://renkulab.io/


## Get notification when finished

* `beepr::beep()`
* Mail notification

### Mail notification

* Package `mail`
    - Very simple
    - No need for mail account
    - Often does not work
* Package `gmailr`
    - More tricky but much better
    - Many resources
        + https://github.com/r-lib/gmailr/blob/master/README.md
        + https://github.com/jennybc/send-email-with-r/blob/master/README.md
        + https://blog.mailtrap.io/r-send-email/

##### Set up `gmailr` {-}

* Enable the Gmail API [here](https://developers.google.com/gmail/api/quickstart/python)
* Download and save credentials
* Run

```{r, eval=FALSE}
library(gmailr)
gm_auth_configure(path = "C:/Users/Reto/scripts/credentials_R_gmailr.json")
gm_auth()
```

##### Send Mails {-}

```{r, eval=FALSE}
library(gmailr)
gm_auth_configure(path = "C:/Users/Reto/scripts/credentials_R_gmailr.json")

alert_mail <- gm_mime() %>%
  gm_to("PUT_A_VALID_EMAIL_ADDRESS_THAT_YOU_CAN_CHECK_HERE") %>%
  gm_from("PUT_THE_GMAIL_ADDRESS_ASSOCIATED_WITH_YOUR_GOOGLE_ACCOUNT_HERE") %>%
  gm_subject("RZ Alert") %>%
  gm_text_body("Computation on IMATH finished")

gm_send_message(alert_mail)
```

##### Allow notifications {-}

Allow notifications for mails with subject RZ Alert

* Thunderbird Add-on `Mailbox Alert`
* Manage message filters
* Add rule that allows notification for specific mails
