# ggplot2

## R Base Graphics vs ggplot2

**R base graphics**

* Based on packages `grapics` and `grDevices`
* Application
  - Graphical parameters are set with the function `par()`
  - High level plot functions create new plot
  - Low level plot functions add elements
* Concept
  - Every plotting function directly prints to the output device
  - What is printed cannot be deleted
  - You can only print on top of an existing plot

**ggplot2**

* Based on package `ggplot2`
* Concept
  - Write a program (saved in a `gg` object) that can later be used to create the plot in one step
  - The whole plot is saved in a single object
  - `ggplot()` creates the `gg` object
  - `+.gg()` adds layers to an existing `gg` object






## Syntax

[Reference page](https://ggplot2.tidyverse.org/reference/)

Overall structure

```{r, eval=FALSE}
ggplot(data = <DATA>) +
  <GEOM_FUNCTION>(mapping = aes(<MAPPINGS>))
```

Detailed structure

```{r, eval=FALSE}
ggplot(data = <DATA>) + 
  <GEOM_FUNCTION>(
     mapping = aes(<MAPPINGS>),
     stat = <STAT>, 
     position = <POSITION>
  ) +
  <COORDINATE_FUNCTION> +
  <FACET_FUNCTION>
```




* `ggplot()` creates a coordinate system where you can add layers
* `geom_point()` scatterplot layer

## Aesthetic mappings `aes`

`aes`: variable => visual properties

* `x` x-axis
* `y` y-axis
* `color` color of points
* `alpha` transparency of points
* `fill` fill color if shape `21:25`
* `shape` shape of points
* `size` size of points
* `stroke` lwd of line arround point

```{block2, type='rmdwarning'}
Only visual properties which map a variable are in `aes()`. Visual properties which affect all points are outsied (example below: col is outside because it affects all points).
```

```{r, out.width='50%'}
require(ggplot2)
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy), col = "blue")
```


* `facet_wrap()`  split plot according to factor
* `facet_grid()`  split plot according to 2 factors


## Goemetric objects `geoms`

`geom`: geometrical object that a plot uses to represent data.

Examples:

* `geom_point()`
* `geom_line()`
* `geom_boxplot()`
* `geom_histogram()`


## Statistical transformations `stat`

`stat`: algorithm to calculate new values for a graph

Check default value of `stat` argument in each `geom`


## Combine base graphics and ggplot2

* Very difficult to combine output of both
  - Usually you have to decide for one of both
* Possible help
  - Package `gridBase` helps to combine grid (also ggplot2) and base graphics output
    + [Vignette](https://cran.r-project.org/web/packages/gridBase/vignettes/gridBase.pdf)
  - `ggthemes::theme_base` allows for ggplot output with the base graphics design