# ggplot

[Reference page](https://ggplot2.tidyverse.org/reference/)

Overall structure

```{r, eval=FALSE}
ggplot(data = <DATA>) +
  <GEOM_FUNCTION>(mapping = aes(<MAPPINGS>))
```

Detailed structure

```{r, eval=FALSE}
ggplot(data = <DATA>) + 
  <GEOM_FUNCTION>(
     mapping = aes(<MAPPINGS>),
     stat = <STAT>, 
     position = <POSITION>
  ) +
  <COORDINATE_FUNCTION> +
  <FACET_FUNCTION>
```




* `ggplot()` creates a coordinate system where you can add layers
* `geom_point()` scatterplot layer

## Aesthetic mappings `aes`

`aes`: variable => visual properties

* `x` x-axis
* `y` y-axis
* `color` color of points
* `alpha` transparency of points
* `fill` fill color if shape `21:25`
* `shape` shape of points
* `size` size of points
* `stroke` lwd of line arround point

```{block2, type='rmdwarning'}
Only visual properties which map a variable are in `aes()`. Visual properties which affect all points are outsied (example below: col is outside because it affects all points).
```

```{r, out.width='50%'}
require(ggplot2)
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy), col = "blue")
```


* `facet_wrap()`  split plot according to factor
* `facet_grid()`  split plot according to 2 factors


## Goemetric objects `geoms`

`geom`: geometrical object that a plot uses to represent data.

Examples:

* `geom_point()`
* `geom_line()`
* `geom_boxplot()`
* `geom_histogram()`


## Statistical transformations `stat`

`stat`: algorithm to calculate new values for a graph

Check default value of `stat` argument in each `geom`