# Text output

## Format

Always and exactly 2 places after comma

```{r}
x <- 0.1234567
format(round(x, 2), nsmall = 2)
```



## New lines

Plot text with line break

```{r, fig.height=2, out.width='50%'}
par(mar = c(0,0,0,0)+0.5)
plot(1, type = "n")
text(1,1, labels = "text \n and more text")
```


## Print text to console

Main functions

* `print()`
* `cat()`
* `message()`
* `warning()`
* `stop()`
* `write()`

See [this excellent post on Stack Overflow](https://stackoverflow.com/a/36700294/6152316)

Details

* `print()`
  - Arguments are not pasted together
  - Elements are numbered like `[1]`
  - Quotation mark around strings
* `cat()`
  - Arguments are pasted together (default: `sep = " "`)
  - You have to force new lines with `\n`
  - Most generic `print.*()` functions use `cat()` or mixture of `print.default()` and `cat()` (see for example `getAnywhere(print.lm)`
* `message()`
  - Arguments are pasted together (default: `sep = ""`)
  - Output not sent to stdout (standard output) but to stderr (standard error)
    + Printed in red color
    + can be suppressed and caught with e.g. `supressMessages()`, `tryCatch()`, ...
  

## Combine Text and variables

* `paste()`
* `cat()`
* `format()`
* `sprintf()`
* `glue::glue()`

```{r, collapse=TRUE}
cat("pi is given by", format(pi, digits = 3))
sprintf("pi is given by %.2f", pi)
glue::glue("pi is given by {pi}")
```

