# Time series {#TimeSeries}


## Data sets
```{r}
data("AirPassengers")     # Air passengers 
data("lynx")              # Lynx Trappings
data("lh")                # Luteinizing Hormone
data("EuStockMarkets")    # Eu Stock Markets
data("co2")               # Mauna Loa CO2
```

## Date and Time

Two important types

* Date
* Date-Time

### Date

* Number of days since 1970-01-01
* `Sys.Date` Get current Date 

### Date-Time

* Number of seconds since 1970-01-01 00:00:00
* POSIXct: number
* POSIXlt: list of character specifying day, hour, min, sec, ...
* `Sys.time` Get current Date-Time
* `as.Date()` coerce
* `strptime()` convert between character and POSIXct


## Time series classes

Object classes which are meant for time series


### ts

* Simple
* Only regularly spaced TS

#### Basic functions

* `ts()` create time-series objects
* `start()` time of first observation (number) 
* `end()` time of last observation (number)
* `frequency()` the number of observations per unit of time
* `deltat()` the fraction of the sampling period between successive observations
* `time()` vector of times at which a time series was sampled
* `window()` subsetting ts objects


```{r, collapse=TRUE}
myts <- ts(c(10,12,14,13,12,10,5,2,15,23,25,27), start=2004, freq=4)
myts
start(myts)
end(myts)
frequency(myts)
deltat(ts)
time(myts)
window(myts, start=2005, end=2006)
plot(myts)
```



### zoo

* Regularly and irregularly spaced TS


### futher classes

* xts
* its
* tseries



## Differencing

* `diff()`
  - `lag = 1` how far back goes the backshift operator $Y=(1-B^p)X$
  - `differences = 1` how many times iteratively apply differencing $Y=(1-B)^dX$
